---
title: "lab_01"
author: "Derek Willis"
date: "2025-01-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# Turn off scientific notation
options(scipen=999)
```

## About this lab

To complete this lab, you need to: \* run existing code as directed (look for **Task**). \* modify existing code as directed (look for **Task**). \* write code in empty codeblocks provided to answer questions included (look for **Q**). \* write out the answer in the form of a complete sentence in the space given (look for **A**). When you are finished, commit changes and push to your personal GitHub repo, then submit the URL to this document on ELMS.

Remember to add any new verbs or concepts to your class reference notebook!

## Load libraries and establish settings

**Task** Run the codeblock below to load the Tidyverse. To run the codeblock, hit the little green play button in the upper right corner of the gray area (the codeblock) below.

```{r}
# Load the tidyverse. If you have not installed the tidyverse already, remove the # from the next line and run it first.  
# install.packages('tidyverse')
library(tidyverse)
```

## Load Data

We're going to load the UMD courses data.

**Task** Run the codeblock below to load the data, which is contained in an RDS file.

```{r}
umd_courses <- read_rds("lab_01.rds")
```

## Answer questions

**Q1** Which department has the highest mean (average) number of seats for its classes? You need to show the department with the highest average first. Which one is the most surprising/interesting to you? **A1** CMNS has the highest average number of seats. I was most surprised by Air Science having a high seat count, as I thought it was a more niche major. 

```{r}
umd_courses |>
  group_by(department) |>
  summarise(
    count_classes = n(),
    total_seats = sum(seats),
    mean_seats = mean(seats),
    median_seats = median(seats)
  ) |>
  arrange(desc(mean_seats))
```

**Q2** Find out the greatest number of times any individual class has been offered in our dataset. Your code must show the greatest number of classes first.
**A2** The greatest number of times any individual class has been offered is 29.

```{r}
umd_courses |>
  group_by(id) |>
  summarise(
    count_class = n()
  ) |>
  arrange(desc(count_class))
```

**Q3** Count the number of courses for each instructor and department. Which instructor has taught the most?
**A3** Jason Kuo, who I've had before! Lovely guy. 

```{r}
umd_courses |>
  group_by(instructors) |>
  summarise(
    count_instructors = n()
  ) |>
  arrange(desc(count_instructors))

umd_courses |>
  group_by(department) |>
  summarise(
    count_dept = n()
  ) |>
  arrange(desc(count_dept))
```

**Q4** Given the results for Q3, what would have made it easier to find the answer? You don't have to know *how* to accomplish this; I want you to describe in words what would make the answer easier to see. How confident are you that the answer you got is the right one, and why? **A4** There are over 15,000 classes without an instructor in the data, which leaves me unconfident that our results are the right one. There is a gulf of missing information that could heavily sway our results away from Jason Kuo. To be fully confident, I would need a complete dataset.

**Q5** Take one of the questions you listed in the final question of the pre_lab and try to answer it here in code. If you don't know how to fully answer it, write what code you can and describe in words what else you would need to do. If you got an answer, is it interesting? **A5** Compared to the last pre-pandemic spring semester, class offerings have increased 5.27% since before the outbreak of COVID-19. This is a minor increase that likely matches the increase in student body size since then, so I don't find this particularly interesting

```{r}
#Answering: Did the COVID-19 pandemic affect the number of courses offered in a semester, or the number of seats in those courses?

#Finding class offerings per semester
umd_courses |>
  group_by(term) |>
  summarise(
    count_classes = n()
  ) |>
  arrange(desc(term))

#Finding the change in class offerings between Spring 2020 and Spring 2025. 
(4893-4648)/4648*100
```
