---
title: "lab_04"
author: "Derek Willis"
date: "2023-10-01"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## About this lab

To complete this lab, you need to: \* run existing code as directed (look for **Task**). \* modify existing code as directed (look for **Task**). \* write code in empty codeblocks provided to answer questions included (look for **Q**). \* write out the answer in the form of a complete sentence in the space given (look for **A**).

When you are finished, commit changes and push to your personal GitHub repo, then submit the URL to this document on ELMS.

## Load libraries and establish settings

You'll need to load three packages for this: the tidyverse, lubridate and janitor.

**Task** load these three packages.

```{r}
# Turn off scientific notation
options(scipen=999)

# load your packages
library(tidyverse)
library(lubridate)
library(janitor)
```

For this lab, we want to investigate spending by Maryland state and local candidates during the 2022 election. For example, we could ask how much money have candidates and committees spent on advertising on Facebook? Which vendors received the most money for media expenses? We have the data, but it's messy - names of payees are spelled differently and capitalized differently - and we need to clean it up. We'll use a combination of RStudio and OpenRefine to do that.

The basic process here is to start in RStudio, export data so that we can use OpenRefine to clean it up, and then bring it back into RStudio to finish our analysis.

## Load Data

You'll need to load one data set, a CSV file of Maryland campaign expenditures located in the data folder called "maryland_expenses.csv"

**Task** Create a codeblock below, then read the data in and assign it to an appropriate variable name. You'll want to clean up the column names and make sure that any date columns are actually date datatypes.

```{r}
# Load the Maryland expenditures data table
maryland_expenses <- read_csv("data/maryland_expenses.csv")
```

## Answer questions

**Q1.** You've been assigned to write about Maryland political committees' use of Facebook advertising. There's a category column in the data that has values for different kinds of spending. One of them is "Media". You'll need to complete the following steps:

1.  Create a dataframe of Media expenses where the name of the payee is not NA.
2.  Then, using `write_csv`, make a CSV file of your media expenses in your data folder. `write_csv()` is similar to `read_csv()` but you provide it with the dataframe you're writing and the file path.
3.  Using OpenRefine, create a project using that CSV file, create a copy of the `payee_name` column called `payee_clean` and then standardize it, focusing on the payees that appear most often. Pay particular attention to variations of Facebook, bearing in mind that it changed its name in 2021: <https://about.fb.com/news/2021/10/facebook-company-is-now-meta/>), but also clean up variations of the most common payees.
4.  When you are done standardizing `payee_clean`, export it as a new CSV file to your data folder, *giving it a different name* and read it back into RStudio as a new dataframe.
5.  Using that `payee_clean` column in your new dataframe, calculate the total amount of Facebook spending by each committee and present the results showing the committee that spent the most first.

**A1.**

```{r}
maryland_expenses_filter <- maryland_expenses |>
  filter(!is.na(`Payee Name`)) |>
  filter(`Expense Category` == "Media")

write_csv(maryland_expenses_filter, "data/maryland_expenses_filter.csv")

maryland_expenses_cleaned <- read_csv("data/maryland_expenses_cleaned.csv")

maryland_expenses_cleaned |>
  filter(payee_clean == "Meta") |>
  group_by(`Committee Name`) |>
  summarise(total = sum(`Amount($)`)) |>
  arrange(desc(total))

```

**Q2.** Let's do some research on the top committees from question 1. Look up the top 10 spenders: if they are candidate committees, what offices were they running for? If they were not candidates, what issues do they seem to care about? Write some sentences answering those questions, then go the Facebook ad library ([https://www.facebook.com/ads/library/?active_status=all&ad_type=political_and_issue_ads&country=US&sort_data\\[direction\\]=desc&sort_data\\[mode\\]=relevancy_monthly_grouped&media_type=all](https://www.facebook.com/ads/library/?active_status=all&ad_type=political_and_issue_ads&country=US&sort_data%5Bdirection%5D=desc&sort_data%5Bmode%5D=relevancy_monthly_grouped&media_type=all){.uri}) and look for ads from the top spender. What do you notice about the the information provided about those ads?

**A2.** Nearly all of the individuals in the top 10 were Democratic Primary candidates throughout Maryland races, ranging from local elections to the gubernatorial race. Few Republicans appeared in the top 10 of my data, and most individuals listed lost their races. CASA in Action was also listed, which backs progressive Democratic candidates in races nationwide. 

The top spender, Ashwani Jain, a candidate for a Maryland congressional seat, pushed ads that showed his position on the War in Gaza and reproductive healthcare. They also advertised his connection to former President Obaam.  

**Q3.** Given the list of top spenders in A1 and the primary results for governor (<https://elections.maryland.gov/elections/2022/primary_results/gen_results_2022_1.html>), who is missing from that Q1 list? Did those campaigns run Facebook ads? Who are the biggest recipients of media expenses by the Democratic primary winner for governor (HINT: that person has two committees, the second one with his running mate)?

**A3.** Both gubernatorial primary victors, Wes Moore and Dan Cox, did not appear to spend heavily on Meta ads during the race. Moore doesn't appear to have spent on Facebook ads, and neither did Cox. Moore's biggest media spend was with SKDK, a campaign management firm that received over six million dollars from Moore's two compaign committees. Foglamp and Apollo Artistry both recieved over ten thousands dollars from the committees, landing in second and third.

```{r}
maryland_expenses_cleaned |>
  filter(payee_clean == "Meta") |>
   filter(`Committee Name` == "Moore  Wes For Maryland" | `Committee Name` == "Moore Miller For Maryland Slate" | `Committee Name` == "Cox  Dan for Governor") |>
  group_by(`Committee Name`) |>
  summarise(total = sum(`Amount($)`)) |>
  arrange(desc(total))

maryland_expenses_cleaned |>
  filter(`Committee Name` == "Moore  Wes For Maryland" | `Committee Name` == "Moore Miller For Maryland Slate") |>
  group_by(payee_clean) |>
  summarise(total = sum(`Amount($)`)) |>
  arrange(desc(total))
```

**Q4.** The media expense data you cleaned also has a standardized purpose column. Write code to generate a list of the committees with the total amount of spending on online advertising. How does this list compare to the one from Q1 about top Facebook spenders? Of the top 5 committees who spent the most on Facebook ads from Q1, are there any that appear to have spent *all* of their online advertising dollars on Facebook?

**A4.** The list is completely different, showing that many campaigns, especially successful ones, spent their digital advertising budgets away from Facebook. Ashwani Jain, the candidate with the highest Facebook spend, appears to have spent his entire digital advertising budget on the platform. The Franchot/Anderson-Walker committee did the same. 

```{r}
maryland_expenses_cleaned |>
  filter(`Expense Purpose` == "Online Advertising") |>
  group_by(`Committee Name`) |>
  summarise(total = sum(`Amount($)`)) |>
  arrange(desc(total))

```

**Q5** Your editor asks you to summarize your findings about spending on Facebook by Maryland political candidates. How much do we know and what don't we know? What's the best story idea you've found here?

**A5** The story that I'd suggest is that Facebook is no longer a priority for successful campaigns in Maryland, as neither Dan Cox nor Wes Moore spent any money on ads on the platform. Those candidates that did largely failed to find electoral success. I'd want to look back at previous campaign spending to see if Larry Hogan's campaigns in 2014 and 2018 spend heavily, or at all, on Facebook ads. It could illustrate a changing media landscape, and a change of priority for political campaign consultants. 
